Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-26 review_article.Rmd
scaffold for discussion

The presentation of these rates considered not only the map base, but also appropriate choices of colour blind friendly colour schemes, and categories of the values to highlight signifcantly different areas.
<!--

@@

The primary display of an atlas is often the choropleth map.
Cruickshank's (1947) as cited by @DMAHP, discusses using visuals as 'formal statistical assessment of the spatial pattern'. 
Atlases can also utilise bar charts, tables and scatterplots as useful alternative displays. 
These displays can present the cancer incidence and mortality rates, as well as the supporting statistics of the atlas. 
When presenting cancer atlases, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses.'
@CPISACA suggests additional materials such as 'good tables, graphs, and explanatory text' support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers. 
The interactivity of modern maps enables supplementary information to be incorporated without cluttering the screen, such as in a tool tip feature. 
The atlases provided additional information to help users understand and interpret the spatial distribution presented. The use of these supports were found in a variety of online map sources.
These additional pieces of information can help to validate, explain or explore relationships between cancer incidence and survival rates.

@@

### Demographics
Demographics include information regarding the age and sex distribution of the areas presented. 
<!-- sex -->
Sex is an important cofactor for cancer atlases. 
As some cancers are sex specific, and others may be found in both males and females, atlases often specify their use of gender in the diplays. 
Atlases can allow for interaction as users can select between displays for males, females or both depending on the type of cancers explored.

### Socioeconomic indicators
Socioeconomic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates and percentages, remoteness and education levels achieved though only a few atlases also explored the impact of rurality on cancer rates. 
The Human Development Index can be used to understand the socioeconomic experience of a community, as can the Income levels as measured by World bank list of economies[@IARC_3]. 
The areas are usually ranked and allocated to quintiles to be presented as categories describing the ranking.

@@

Non-Contiguous cartograms succeed in maintaing the shape of the areas presented.
Each area stays in a similar position to their location on a choropleth map. The choropleth map base is often also plotted as a comparison point to highlight the change in area. The addition is the gap between areas, created as each individual area shrinks or grows according the associated value of the statistic. @NAC discusses creation of these maps, the significance of the empty areas left between the geographic boundaries and the new shape, and the 'degree of difference from the original map that is the real message' of these displays.

@@

He answers this with his implementation of maps created with 'the simplest of all shapes'. He acknowledged the sophistication of contiguous cartograms but critiqued their 'very complex shapes'.
Circular cartograms use the same simple shapes for every region represented, and size the shapes according to the statistic represented or the population for a base map. This familar shape may be more effective for understanding the spatial distribution than contiguous cartograms, as the 'nonsense' shapes used have 'no meaning' after distortions are applied @NISCC.
To produce a compelling map, a gravity model is applied to avoid overlaps, and keep spatial relationships with neighbouring areas over many iterations. This implementation can work for up to 'one hundred thousand' areas.

@@

The groundwork for the simplistic approach had been laid in the mid 1930's be Raisz. @TFYCC quotes the official definition of Value-Area Cartograms, the simplistic displays which represent each area as a single rectangle, sized according to the value of the statistic.
This rectangluar display also allows for tiling, where geographic neighbours also share rectangle borders.

# A critique of mapping methods

@@

>>>>

The presentation of these rates considered not only the map base, but also appropriate choices of colour blind friendly colour schemes, and categories of the values to highlight significantly different areas.
<!--

@@

The primary display of an atlas is often the choropleth map.
Cruickshank's (1947) as cited by @DMAHP, discusses using visuals as 'formal statistical assessment of the spatial pattern'. 
Atlases can also utilise bar charts, tables and scatter-plots as useful alternative displays. 
These displays can present the cancer incidence and mortality rates, as well as the supporting statistics of the atlas. 
When presenting cancer atlases, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses.'
@CPISACA suggests additional materials such as 'good tables, graphs, and explanatory text' support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers. 
The interactivity of modern maps enables supplementary information to be incorporated without cluttering the screen, such as in a tool tip feature. 
The atlases provided additional information to help users understand and interpret the spatial distribution presented. The use of these supports were found in a variety of online map sources.
These additional pieces of information can help to validate, explain or explore relationships between cancer incidence and survival rates.

@@

### Demographics
Demographics include information regarding the age and sex distribution of the areas presented. 
<!-- sex -->
Sex is an important cofactor for cancer atlases. 
As some cancers are sex specific, and others may be found in both males and females, atlases often specify their use of gender in the displays. 
Atlases can allow for interaction as users can select between displays for males, females or both depending on the type of cancers explored.

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates and percentages, remoteness and education levels achieved though only a few atlases also explored the impact of rurality on cancer rates. 
The Human Development Index can be used to understand the socio-economic experience of a community, as can the Income levels as measured by World bank list of economies[@IARC_3]. 
The areas are usually ranked and allocated to quintiles to be presented as categories describing the ranking.

@@

Non-Contiguous cartograms succeed in maintaining the shape of the areas presented.
Each area stays in a similar position to their location on a choropleth map. The choropleth map base is often also plotted as a comparison point to highlight the change in area. The addition is the gap between areas, created as each individual area shrinks or grows according the associated value of the statistic. @NAC discusses creation of these maps, the significance of the empty areas left between the geographic boundaries and the new shape, and the 'degree of difference from the original map that is the real message' of these displays.

@@

He answers this with his implementation of maps created with 'the simplest of all shapes'. He acknowledged the sophistication of contiguous cartograms but critiqued their 'very complex shapes'.
Circular cartograms use the same simple shapes for every region represented, and size the shapes according to the statistic represented or the population for a base map. This familiar shape may be more effective for understanding the spatial distribution than contiguous cartograms, as the 'nonsense' shapes used have 'no meaning' after distortions are applied @NISCC.
To produce a compelling map, a gravity model is applied to avoid overlaps, and keep spatial relationships with neighbouring areas over many iterations. This implementation can work for up to 'one hundred thousand' areas.

@@

The groundwork for the simplistic approach had been laid in the mid 1930's be Raisz. @TFYCC quotes the official definition of Value-Area Cartograms, the simplistic displays which represent each area as a single rectangle, sized according to the value of the statistic.
This rectangular display also allows for tiling, where geographic neighbours also share rectangle borders.

# A discussion of mapping methods

@@

<!-- Choropleths, true topology, geographic distance, good for showing geographic importance -->
Visualisation properties have been explored across several disease mapping displays.
Choropleth displays utilise more traditional cartographic methods, this means they are usually true to the topography.
Tobler's work highlights the need for consideration of the transformation of the globe to a 2D space.
When there is a relationship between the cancer type being mapped and geographical features, a choropleth map would pair this information visually.
These maps are best used when the geography is of importance to the data in the display.
Users can identify the areas relevant to them by the familiar boundaries, and can easily contrast the experience of geographic neighbours.

<!-- Cartograms, sacrifice geography, focus on the outliers/ relationships in data space -->
Pairing geography and cancer data was historically the work of epidemiologists, however it did not require them to be the map designer.
This was expected as cartographers were the creators of geographic displays, and epidemiologists utilised the maps that had been prepared in advance.
This disconnect did not allow the map to be prepared with the data in mind.
Cartogram methods began exploring the concept of diagrammatic cartograms, displays that highlighted the importance of data, as well as the geography.
This approach began a shift in the focus and responsibility of map creators.
Cartogram displays began incorporating the statistics and population into the design of the display. 
Shifting and sometimes sacrificing familiar boundaries to draw attention to outliers in the data space or population density.
The difference between the familiar map display helps highlight the impact of the disease on communities.

<!-- Alternatives, remove geographic shapes, keep spatial relationship, focus on the outliers/ relationships in population space -->
By removing the unexpected boundaries and shapes, the spatial distribution is the primary concept communicated by the display.
Areas that are geographically close will maintain connectedness in some way, but the population and the statistic will take precedence.
These alternatives are especially helpful for data aggregations where administrative boundaries break populations into groups.
The experience of each collective may be worth considering as their experiences may be similar in terms of the services and facilities they share.
Population affects are highlighted by these displays, as this display allows a more equitable view of each community, and does not minimise those that operate on a smaller geographic scale.
